---
title: "Comparison of different methods for calculationg weights in epiregulon with the use of the simulated data"
author: "Tomasz Włodarczyk"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
# Introduction
This notebook compares across all options for weight computation which are available
in the epiregulon package. As a ground truth the data simulated by scMultiSim package is used. 

# Highlights

- GRN is composed of 5000 target genes and 500 transcription factors

# GRN preparation

```{r}
Sys.setenv("R_MAX_VSIZE" = 32*2^30)
```

```{r child = "GRN_preparation.Rmd", cache = TRUE}
```

```{r child = "basic_simulation.Rmd", cache = TRUE}
```

# Comparison of weight methods based on the true counts (unnormalized)

```{r cache = TRUE}
library(epiregulon)
library(epiregulon.benchmark)
input_objects <- processSimResults(sim_res)

# skip pruning since all connections are true
# add weights using wilcoxon test
regulon.w.wilcoxon <- addWeights(input_objects$regulon, peakMatrix = input_objects$peakMatrix,
                        expMatrix = input_objects$geneExpMatrix,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "wilcoxon",
                        clusters = NULL)

regulon.w.logFC <- addWeights(input_objects$regulon, peakMatrix = input_objects$peakMatrix,
                        expMatrix = input_objects$geneExpMatrix,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "logFC",
                        clusters = NULL)

regulon.w.corr <- addWeights(input_objects$regulon, peakMatrix = input_objects$peakMatrix,
                        expMatrix = input_objects$geneExpMatrix,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "corr",
                        clusters = input_objects$geneExpMatrix$label)

regulon.w.lmfit <- addWeights(input_objects$regulon, peakMatrix = input_objects$peakMatrix,
                        expMatrix = input_objects$geneExpMatrix,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "lmfit",
                        clusters = input_objects$geneExpMatrix$label)

regulon.w.MI <- addWeights(input_objects$regulon, peakMatrix = input_objects$peakMatrix,
                        expMatrix = input_objects$geneExpMatrix,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "MI",
                        clusters = input_objects$geneExpMatrix$label)


score.combine.wilcoxon <- calculateActivity(regulon = regulon.w.wilcoxon, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix,
                                   exp_assay = "counts"
                                   ) 

score.combine.logFC <- calculateActivity(regulon = regulon.w.logFC, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix,
                                   exp_assay = "counts"
                                   ) 

score.combine.lmfit <- calculateActivity(regulon = regulon.w.lmfit, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix,
                                   exp_assay = "counts"
                                   ) 

score.combine.corr <- calculateActivity(regulon = regulon.w.corr, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix,
                                   exp_assay = "counts"
                                   ) 

score.combine.MI <- calculateActivity(regulon = regulon.w.MI, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix,
                                   exp_assay = "counts"
                                   ) 

score_matrices <-list(score.combine.wilcoxon, score.combine.logFC, score.combine.lmfit,
                      score.combine.corr, score.combine.MI)
methods <- c("wilcoxon", "logFC", "lmfit", "corr", "MI")

```

```{r cache = TRUE}
df <- data.frame()
for(i in seq_along(score_matrices)){
    df <- rbind(df, data.frame(value = assessRankAccuracy(GRN = GRN, 
                                                          sim_options = options_, 
                                                          basic_sim_res = sim_res, 
                                                          activities_obs = score_matrices[[i]]),
                               method = methods[i]))
}

boxplot(df$value~df$method, xlab = "Weight method", ylab = "Rankings mismatch")
```


# Comparison of weight methods based on the observed counts (unnormalized)

```{r cache = TRUE}
regulon.w.wilcoxon <- addWeights(input_objects$regulon, peakMatix_obs = input_objects$peakMatix_obs,
                        expMatrix = input_objects$geneExpMatrix_obs,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "wilcoxon",
                        clusters = NULL)

regulon.w.logFC <- addWeights(input_objects$regulon, peakMatix_obs = input_objects$peakMatix_obs,
                        expMatrix = input_objects$geneExpMatrix_obs,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "logFC",
                        clusters = NULL)

regulon.w.corr <- addWeights(input_objects$regulon, peakMatix_obs = input_objects$peakMatix_obs,
                        expMatrix = input_objects$geneExpMatrix_obs,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "corr",
                        clusters = input_objects$geneExpMatrix_obs$label)

regulon.w.lmfit <- addWeights(input_objects$regulon, peakMatix_obs = input_objects$peakMatix_obs,
                        expMatrix = input_objects$geneExpMatrix_obs,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "lmfit",
                        clusters = input_objects$geneExpMatrix_obs$label)

regulon.w.MI <- addWeights(input_objects$regulon, peakMatix_obs = input_objects$peakMatix_obs,
                        expMatrix = input_objects$geneExpMatrix_obs,
                        exp_assay = "counts",
                        peak_assay = "peak",
                        method = "MI",
                        clusters = input_objects$geneExpMatrix_obs$label)


score.combine.wilcoxon <- calculateActivity(regulon = regulon.w.wilcoxon, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix_obs,
                                   exp_assay = "counts"
                                   ) 

score.combine.logFC <- calculateActivity(regulon = regulon.w.logFC, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix_obs,
                                   exp_assay = "counts"
                                   ) 

score.combine.lmfit <- calculateActivity(regulon = regulon.w.lmfit, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix_obs,
                                   exp_assay = "counts"
                                   ) 

score.combine.corr <- calculateActivity(regulon = regulon.w.corr, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix_obs,
                                   exp_assay = "counts"
                                   ) 

score.combine.MI <- calculateActivity(regulon = regulon.w.MI, mode = "weight", 
                                   method = "weightedMean",
                                   expMatrix = input_objects$geneExpMatrix_obs,
                                   exp_assay = "counts"
                                   ) 

score_matrices <-list(score.combine.wilcoxon, score.combine.logFC, score.combine.lmfit,
                      score.combine.corr, score.combine.MI)
methods <- c("wilcoxon", "logFC", "lmfit", "corr", "MI")

```

```{r cache = TRUE}
df <- data.frame()
for(i in seq_along(score_matrices)){
    df <- rbind(df, data.frame(value = assessRankAccuracy(GRN = GRN, 
                                                          sim_options = options_, 
                                                          basic_sim_res = sim_res, 
                                                          activities_obs = score_matrices[[i]]),
                               method = methods[i]))
}

boxplot(df$value~df$method, xlab = "Weight method", ylab = "Rankings mismatch")
```
