---
title: "Comparison of different methods for calculationg weights in epiregulon with the use of NGS4557 dataset"
author: "Tomasz Włodarczyk"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
# Introduction
This notebook compares across all options for weight computation which are available
in the epiregulon package. As a ground truth NGS4557 dataset is used coming from the study on
the effect of drug treatment on the gene expression of LNCaP and VCaP cell lines. The results are
presented in the form of curves showing the relation between false and true positive rates.

# Highlights

- clusters in methods "corr", "lmfit" and "MI" defined by 'Clusters_Combined' column in the gene expression SCE
- cell line specific weights used for calculation of activity when 'wilcoxon' and 'logFC' methods are used 
- motif scores added
- filtering on motif scores

# Regulon preparation

```{r Epiregulon, child = system.file("reports/pruned_regulon_AR.Rmd", package = "epiregulon.benchmark"),cache=TRUE, cache.lazy==FALSE}
```


# Preparation of the data frame to be plotted

```{r weights, cache = TRUE, cache.lazy = FALSE}
library(ArchR)
peakMatrix <- peakMatrix_base
GeneExpressionMatrix <- GeneExpressionMatrix_base

library(epiregulon.benchmark)
group_combinations_1 <- expand.grid(c(params$cell_line), c("ARV110", "Enza"), "DMSO", "down", "AR")
group_combinations_2 <- expand.grid(c(params$cell_line), c("A9690"), "DMSO", "down", c("SMARCA2", "SMARCA4"))
group_combinations <- rbind(group_combinations_1, group_combinations_2)
colnames(group_combinations) <- c("cell_line", "experimental_treatment", "control_treatment", "effect_direction", "transcription_factor")

plot_data <- data.frame()
```

```{r plots, cache = TRUE, cache.lazy = FALSE, results = "hide", , message=FALSE, warning=FALSE}
arg.list <- list()
arg.list$peakMatrix = peakMatrix
arg.list$peak_assay = "counts"
arg.list$exp_assay = "normalizedCounts"
weight_clusters <- as.list(c(corr = "Clusters_Combined", MI = "Clusters_Combined", lmfit = "Clusters_Combined", logFC = NULL, wilcoxon = NULL))
activity_clusters <- c(corr = NA, MI = NA, lmfit = NA, logFC = NA, wilcoxon = NA)
for (method in  c("corr", "MI", "lmfit","logFC","wilcoxon")){
    arg.list$method = method
    partial_data <- prepare_plot_data(regulon = pruned.regulon,
                                      weight.args = arg.list,
                                      group_combinations = group_combinations,
                                      geneExprMatrix.sce = GeneExpressionMatrix,
                                      weight_clusters = weight_clusters[[method]],
                                      activity_clusters = activity_clusters[method],
                                      motif_score = FALSE,
                                      path_to_archR_project = "/gstore/project/ar_ligands/AR/scRNAseq/NGS4557_BRG1i/OUTPUT/gp.sa.archr/")
    plot_data <- rbind(plot_data, partial_data)
}

```

```{r AUC, cache = TRUE, cache.lazy = FALSE}
library(dplyr)
library(ggplot2)

ggplot(plot_data, aes(x=FPR, y=TPR))+
    geom_line(aes(linetype = cell_line, color = tf))+
    facet_grid(treatment~method)+
    xlab("False positive rate")+
    ylab("True positive rate")+
    ggtitle("ROC curves for AR dataset")

AUC_data <- plot_data %>% group_by(cell_line, treatment, method, tf) %>%
    group_map(function(x,y) cbind(y, data.frame(AUC = calculate_AUC(x$FPR, x$TPR))))

AUC_data <- do.call(rbind, AUC_data)

print(AUC_data)

GRN_list <- list()
regulon <- addWeights(regulon = pruned.regulon, expMatrix = GeneExpressionMatrix, peakMatrix = peakMatrix, exp_assay = "normalizedCounts", peak_assay = "counts", method = "wilcoxon", clusters = GeneExpressionMatrix$TEST_ARTICLE,
exp_cutoff = NULL)
GRN_list$Epiregulon <- regulon
```

```{r FigR, child = system.file("reports/FigR_AR.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy==FALSE}
```

```{r Pando, child = system.file("reports/Pando_AR.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy==FALSE}
```


```{r cellOracle, child = system.file("reports/cellOracle_AR.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy==FALSE}
```

```{r GRaNIE, child = system.file("reports/GRaNIE_AR.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy==FALSE}
```

```{r Scenic_plus, child = system.file("reports/Scenic_plus_AR.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy==FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["Pando"]] <- Seurat_obj 
GRN_list[["cellOracle"]] <- links
GRN_list[["FigR"]] <- FigR_GRN
GRN_list[["GRaNIE"]] <- GRN_connections.all
library(epiregulon)
library(epiregulon.benchmark)
GeneExpressionMatrix  <- GeneExpressionMatrix_base
SummarizedExperiment::colData(GeneExpressionMatrix)$cluster_cellOracle <- louvain_clusters

matrices_list <- list()
if(length(assays(GeneExpressionMatrix))!= 1) stop("GeneExpressionMatrix should contain one assay")
for(i in seq_along(GRN_list)){
    GRN <- GRN_list[[i]]
    method <- names(GRN_list)[i]
    activity.matrix <- get_activity_matrix(GRN = GRN,
                                           method = method,
                                           geneExprMatrix.sce = GeneExpressionMatrix,
                                           tfs = c("AR", "SMARCA2", "SMARCA4"),
                                           exp_assay = names(assays(GeneExpressionMatrix))[1])
  matrices_list[[i]] <- activity.matrix 
}
names(matrices_list) <- names(GRN_list)
```

```{r}
gene_based <- scenic_plus_res[[1]]
region_based <- scenic_plus_res[[2]]
sample_id <- gsub("(.*___)(.*)", "\\2",region_based$Cell)
barcode <- gsub("(.*)(___.*)", "\\1",region_based$Cell)
region_based_cell_names <- paste0(sample_id, "#", barcode)

sample_id <- gsub("(.*___)(.*)", "\\2",gene_based$Cell)
barcode <- gsub("(.*)(___.*)", "\\1",gene_based$Cell)
gene_based_cell_names <- paste0(sample_id, "#", barcode)

# first column stores old cell names
region_based_genes <- colnames(region_based)[-1]
gene_based_genes <- colnames(gene_based)[-1]
# convert from character to numeric
region_based <- apply(as.matrix(region_based[,2:ncol(region_based)]),2,as.numeric)
region_based <- t(region_based)
colnames(region_based) <- region_based_cell_names
rownames(region_based) <- region_based_genes
gene_based <- apply(as.matrix(gene_based[,2:ncol(gene_based)]),2,as.numeric)
gene_based <- t(gene_based)
colnames(gene_based) <- gene_based_cell_names
rownames(gene_based) <- gene_based_genes
region_based_pos <- region_based[grep("_\\+_", rownames(region_based)),]
region_based_neg <- region_based[grep("_-_", rownames(region_based)),]
rownames(region_based_pos) <- gsub("(^.*)(_\\+_.*)","\\1",rownames(region_based_pos))
rownames(region_based_pos) <- gsub("_extended","",rownames(region_based_pos))
gene_based_pos <- gene_based[grep("_\\+_", rownames(gene_based)),]
gene_based_neg <- gene_based[grep("_-_", rownames(gene_based)),]
rownames(gene_based_pos) <- gsub("(^.*)(_\\+_.*)","\\1",rownames(gene_based_pos))
rownames(gene_based_pos) <- gsub("_extended","",rownames(gene_based_pos))

matrices_list[["scenic_plus_pos_gene"]] <- gene_based_pos
matrices_list[["scenic_plus_neg_gene"]] <- gene_based_neg
matrices_list[["scenic_plus_pos_region"]] <- region_based_pos
matrices_list[["scenic_plus_neg_region"]] <- region_based_neg
```

```{r cache = FALSE, cache.lazy=FALSE}
library(SingleCellExperiment)
library(epiregulon.benchmark)
AUC_table <- data.frame()
for (i in seq_len(nrow(group_combinations))){
    selected_row <- group_combinations[i,,drop = FALSE]
    attach(selected_row)
    tf <- transcription_factor
    main_plot = TRUE
    positive_elements_label <- grep("DMSO", unique(GeneExpressionMatrix$TEST_ARTICLE), value = TRUE)
    negative_elements_label <- grep(experimental_treatment, unique(GeneExpressionMatrix$TEST_ARTICLE), value = TRUE)
    for(i in seq_along(matrices_list)){
        activity.matrix <- matrices_list[[i]]
        if(!tf %in% rownames(activity.matrix)) next
        labels <- colData(GeneExpressionMatrix[,colnames(activity.matrix)])$TEST_ARTICLE
        if(main_plot) {
                AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                              tf = tf,
                              labels = labels,
                              add_plot = FALSE,
                              positive_elements_label = positive_elements_label,
                              negative_elements_label = negative_elements_label,
                              main = sprintf("AR dataset, cell line: %s\ntf: %s", cell_line, tf))
                main_plot <- FALSE
        }
        else{
            AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                                  tf = tf,
                                  labels = labels,
                                  add_plot = TRUE,
                                  positive_elements_label = positive_elements_label,
                                  negative_elements_label = negative_elements_label)
        }
        AUC_table <- rbind(AUC_table, data.frame(AUC = AUC, tf = tf, method = names(matrices_list)[i]))
    }
}
print(AUC_table)
```


```{r}
print(sessionInfo())
```

