---
title: "Comparison of different methods for calculationg weights in epiregulon with the use of reprogram-seq dataset"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
params:
    treatments: list(GATA6 = NULL, NKX2.1 = NULL)
    cluster_column: "Clusters"
    data_file_paths: ""
    temp_dir: ""
    work_dir: ""
    n_cpu: 4
    group_variable: "HTO"
    GRaNIE_tfbs: ""
    save_path: ""
---

```{r setup, include=TRUE, cache = TRUE, cache.lazy=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
# Introduction
This report presents comparison across all method options for weight computation which are available
in the `epiregulon` package. As a ground truth the reprogram-seq dataset is used. It was generated as a result of the studies in which cells were modified to overexpress a selected transcription factor. The results are
presented in the form of curves showing the relation between false and true positive rates.

# Regulon preparation
```{r child = system.file("reports/get_matrices_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy = FALSE}
```

```{r child = system.file("reports/pruned_regulon_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy=FALSE}
```

# Preparation of the data frame to be plotted

```{r cache = TRUE, cache.lazy=FALSE}
library(epiregulon.benchmark)
group_combinations <- expand.grid("", names(treatments), NA, "up", "")
colnames(group_combinations) <- c("cell_line", "experimental_treatment", "control_treatment", "effect_direction", "transcription_factor")
control_treatment <- c()
for(i in seq_along(treatments)){
    if(is.null(treatments[[i]]))
        control_treatment[i] <- get_complementary_names(names(treatments)[i], unique(GeneExpressionMatrix$hash_assignment))
    else
        control_treatment[i] <- paste0(params$treatments[[i]], collapse = "|")
}

group_combinations$control_treatment <- control_treatment
group_combinations$transcription_factor <- unlist(lapply(group_combinations$experimental_treatment, function(x) map_treatment_to_tf(x, regulon)))
plot_data <- data.frame()
```


```{r results = "hide", cache = TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE}
arg.list <- list()
arg.list$peakMatrix = PeakMatrix
arg.list$peak_assay = "counts"
weight_clusters  <- as.list(setNames(rep(params$cluster_column, 3), c("corr", "MI", "lmfit")))
for (method in  c("corr", "MI", "lmfit","logFC","wilcoxon")){
    arg.list$method = method
    partial_data <- prepare_plot_data(regulon = regulon,
                                      weight.args = arg.list,
                                      group_combinations = group_combinations,
                                      geneExprMatrix.sce = GeneExpressionMatrix,
                                      weight_clusters = weight_clusters[[method]],
                                      activity_clusters = NA)
    plot_data <- rbind(plot_data, partial_data)
}

```

```{r cache = TRUE, cache.lazy=FALSE}
library(dplyr)
library(ggplot2)

ggplot(plot_data, aes(x=FPR, y=TPR))+
    geom_line()+
    facet_grid(treatment~method)+
    xlab("False positive rate")+
    ylab("True positive rate")+
    ggtitle("ROC curves for reprogram-seq dataset")


AUC_data <- plot_data %>% group_by(cell_line, treatment, method, tf) %>%
    group_map(function(x,y) cbind(y, data.frame(AUC = calculate_AUC(x$FPR, x$TPR))))
AUC_data <- do.call(rbind, AUC_data)

print(AUC_data)
```
```{r echo = FALSE, cache.lazy=FALSE}
if(!is.null(params$output_file_summary)) write.csv(AUC_data, params$output_file_summary)
```
```{r cache = TRUE, cache.lazy=FALSE}
GRN_list <- list()
weight.args <- list(clusters = as.vector(GeneExpressionMatrix[[params$cluster_column]]), regulon = regulon,
                    expMatrix = GeneExpressionMatrix, peakMatrix = PeakMatrix, peak_assay = "counts",
                    method = "corr")
# epiregulon workflow
regulon.w <- do.call(addWeights, weight.args)
GRN_list[["Epiregulon"]] <- regulon.w
GeneExpressionMatrix_copy <- GeneExpressionMatrix
```

```{r child = system.file("reports/Pando_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE}
Seurat_obj  <- Seurat::RenameCells(Seurat_obj, new.names = paste0("reprogram#", colnames(Seurat_obj))) 
GRN_list[["Pando"]] <- Seurat_obj 
```


```{r child = system.file("reports/cellOracle_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["cellOracle"]] <- links
```

```{r child = system.file("reports/FigR_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE,message=FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["FigR"]] <- FigR_GRN
```

```{r child = system.file("reports/GRaNIE_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE,message=FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["GRaNIE"]] <- GRN_connections.all
```

```{r child = system.file("reports/Scenic_plus_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE}
```

```{r cache = TRUE, cache.lazy=FALSE, message=FALSE}
GeneExpressionMatrix <- GeneExpressionMatrix_copy
louvain_clusters$barcode = gsub("(.*)(___)(.*)", paste0("\\3","#","\\1"), louvain_clusters$barcode)
louvain_clusters$barcode <- gsub("\\.","-",louvain_clusters$barcode)
SummarizedExperiment::colData(GeneExpressionMatrix)$cluster_cellOracle <- as.character(louvain_clusters$cluster[match(louvain_clusters$barcode, colnames(GeneExpressionMatrix))])
matrices_list <- list()
for(i in seq_along(GRN_list)){
    GRN <- GRN_list[[i]]
    method <- names(GRN_list)[i]
    activity.matrix <- get_activity_matrix(GRN = GRN,
                                           method = method,
                                           geneExprMatrix.sce = GeneExpressionMatrix,
                                           tfs = c("GATA6", "NKX2-1", "FOXA2"))
  matrices_list[[i]] <- activity.matrix 
}

names(matrices_list) <- names(GRN_list)
region_based <- scenic_plus_res$region_based
gene_based <- scenic_plus_res$gene_based
region_based_cell_names <- gsub("\\.","-",(gsub("___reprogram","",region_based$Cell)))
region_based_cell_names <- paste0("reprogram#",region_based_cell_names)
gene_based_cell_names <- gsub("\\.","-",(gsub("___reprogram","",gene_based$Cell)))
gene_based_cell_names <- paste0("reprogram#",gene_based_cell_names)
# first column stores old cell names
region_based_genes <- colnames(region_based)[-1]
gene_based_genes <- colnames(gene_based)[-1]
# convert from character to numeric
region_based <- apply(as.matrix(region_based[,2:ncol(region_based)]),2,as.numeric)
region_based <- t(region_based)
colnames(region_based) <- region_based_cell_names
rownames(region_based) <- region_based_genes
gene_based <- apply(as.matrix(gene_based[,2:ncol(gene_based)]),2,as.numeric)
gene_based <- t(gene_based)
colnames(gene_based) <- gene_based_cell_names
rownames(gene_based) <- gene_based_genes
region_based_pos <- region_based[grep("_act_", rownames(region_based)),]
region_based_neg <- region_based[grep("_supp_", rownames(region_based)),]
rownames(region_based_pos) <- gsub("(^.*)(_act_.*)","\\1",rownames(region_based_pos))
rownames(region_based_pos) <- gsub("_extended","",rownames(region_based_pos))
gene_based_pos <- gene_based[grep("_act_", rownames(gene_based)),]
gene_based_neg <- gene_based[grep("_supp_", rownames(gene_based)),]
rownames(gene_based_pos) <- gsub("(^.*)(_act_.*)","\\1",rownames(gene_based_pos))
rownames(gene_based_pos) <- gsub("_extended","",rownames(gene_based_pos))

matrices_list[["scenic_plus_pos_gene"]] <- gene_based_pos
matrices_list[["scenic_plus_neg_gene"]] <- gene_based_neg
matrices_list[["scenic_plus_pos_region"]] <- region_based_pos
matrices_list[["scenic_plus_neg_region"]] <- region_based_neg
```


```{r cache = FALSE, cache.lazy=FALSE, message=FALSE}
library(SingleCellExperiment)
library(epiregulon.benchmark)
AUC_table <- data.frame()
colors <- c("black", "blue", "red", "green1", "purple", "grey", "orange", "bisque", "green4","tan4")[1:length(matrices_list)]
colors <- setNames(colors, names(matrices_list))
for (tf in c("GATA6", "NKX2-1", "FOXA2")){
    hashtag_pattern <- gsub("-", ".", tf)
    main_plot = TRUE
    positive_elements_label <- grep(hashtag_pattern, unique(GeneExpressionMatrix$hash_assignment), value = TRUE)
    negative_elements_label <- grep(hashtag_pattern, unique(GeneExpressionMatrix$hash_assignment), value = TRUE, invert = TRUE)
    for(i in seq_along(matrices_list)){
        activity.matrix <- matrices_list[[i]]
        if(!tf %in% rownames(activity.matrix)) next
        labels <- colData(GeneExpressionMatrix[,colnames(activity.matrix)])$hash_assignment
        if(main_plot) {
                AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                              tf = tf,
                              labels = labels,
                              add_plot = FALSE,
                              positive_elements_label = positive_elements_label,
                              negative_elements_label = negative_elements_label,
                              main = sprintf("Reprogram-seq dataset\n%s", tf),
                              col = colors[i]
                              )
                main_plot <- FALSE
        }
        else{
            AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                                  tf = tf,
                                  labels = labels,
                                  add_plot = TRUE,
                                  positive_elements_label = positive_elements_label,
                                  negative_elements_label = negative_elements_label,
                                  col = colors[i])
        }
        AUC_table <- rbind(AUC_table, data.frame(AUC = AUC, tf = tf, method = names(matrices_list)[i]))
    }
    legend("bottomright", names(colors), col = colors, lty = rep(1, length(colors)))
}
print(AUC_table)
```

