---
title: "Comparison of different methods for calculationg weights in epiregulon with the use of reprogram-seq dataset"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
params:
    treatments: list(GATA6 = NULL, NKX2.1 = NULL, FOXA2 = NULL)
    cluster_column: "Clusters"
    data_file_paths: ""
    temp_dir: ""
    work_dir: ""
    n_cpu: 4
    group_variable: "HTO"
    GRaNIE_tfbs: ""
    save_path: ""
    packages: "c('Epiregulon', 'Pando', 'cellOracle', 'scenicplus', 'GRaNIE', 'FigR')"
    save_results: TRUE
    path_to_archR_project: ""
    motif_db_dir: ""
    
---

```{r setup, include=TRUE, cache = TRUE, cache.lazy=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
# Introduction
This report presents comparison across all method options for weight computation which are available
in the `epiregulon` package. As a ground truth the reprogram-seq dataset is used. It was generated as a result of the studies in which cells were modified to overexpress a selected transcription factor. The results are
presented in the form of curves showing the relation between false and true positive rates.

# Regulon preparation
```{r child = system.file("reports/get_matrices_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy = FALSE}
```

```{r child = system.file("reports/pruned_regulon_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache=TRUE, cache.lazy=FALSE}
```

```{r echo=FALSE}
treatments <- eval(parse(text = enquote(params$treatments)))
benchmarked_packages <- eval(parse(text = enquote(params$packages)))
save_path <- params$save_path
if(params$save_results && save_path == ""){
    if(!dir.exists(file.path(getwd(), "benchmark_output")))
        dir.create(file.path(getwd(), "benchmark_output"))
    if(!dir.exists(file.path(getwd(), "benchmark_output", "plots")))
        dir.create(file.path(getwd(), "benchmark_output", "plots"))
    save_path <- file.path(getwd(), "benchmark_output")
}
if(!dir.exists(file.path(save_path, "plots")))
    dir.create(file.path(save_path, "plots"))
```

# Preparation of the data frame to be plotted

```{r cache = TRUE, cache.lazy=FALSE}
library(epiregulon.benchmark)
group_combinations <- expand.grid(NA, names(treatments), NA, "up", "")
colnames(group_combinations) <- c("cell_line", "experimental_treatment", "control_treatment", "effect_direction", "transcription_factor")
control_treatment <- c()
for(i in seq_along(treatments)){
    if(is.null(treatments[[i]]))
        control_treatment[i] <- get_complementary_names(names(treatments)[i], unique(GeneExpressionMatrix$hash_assignment))
    else
        control_treatment[i] <- paste0(params$treatments[[i]], collapse = "|")
}

group_combinations$control_treatment <- control_treatment
group_combinations$transcription_factor <- unlist(lapply(group_combinations$experimental_treatment, function(x) map_treatment_to_tf(x, regulon)))
plot_data <- data.frame()
```


```{r results = "hide", cache = TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE}
arg.list <- list()
arg.list$peakMatrix = PeakMatrix
arg.list$peak_assay = "counts"
arg.list$tf_re.merge = TRUE
arg.list$exp_assay = "normalizedCounts"
weight_clusters  <- as.list(setNames(rep(params$cluster_column, 3), c("corr", "MI", "lmfit")))
save.image(".RData")
for (method in  c("corr", "MI", "lmfit","logFC","wilcoxon")){
    arg.list$method = method
    partial_data <- prepare_plot_data(regulon = regulon,
                                      weight.args = arg.list,
                                      group_combinations = group_combinations,
                                      geneExprMatrix.sce = GeneExpressionMatrix,
                                      weight_clusters = weight_clusters[[method]],
                                      activity_clusters = NA,
                                      motif_score = FALSE)
    plot_data <- rbind(plot_data, partial_data)
}

```

```{r weight_plots, cache = TRUE, cache.lazy=FALSE}
library(dplyr)
library(ggplot2)

ggplot(plot_data, aes(x=FPR, y=TPR))+
    geom_line(linewidth = 0.2)+
    facet_grid(tf~method)+
    xlab("False positive rate")+
    ylab("True positive rate")+
    ggtitle("ROC curves for reprogram-seq dataset")+
    theme(panel.border = element_rect(color = "black", fill = NA),
          panel.background = element_rect(fill = "white", colour="white"),
          plot.title = element_text(hjust = 0.5))

if(params$save_results)
    ggsave(file.path(save_path, "plots", "Reprogram_weight_methods.pdf"), width = 10, height = 5.5)

AUC_data <- plot_data %>% group_by(method, tf) %>%
    group_map(function(x,y) cbind(y, data.frame(AUC = calculate_AUC(x$FPR, x$TPR))))

AUC_data <- do.call(rbind, AUC_data)

print(AUC_data)

if(params$save_results){
    write.csv(AUC_data, file.path(save_path, "AUC_reprogram_by_weight_method.csv"))
    write.csv(plot_data, file.path(save_path, "plot_data_reprogram_by_weight_method.csv"))
}

# save copy of GeneExpressionMatrix since it might be changed in the following workflows
GeneExpressionMatrix_copy <- GeneExpressionMatrix
```
```{r echo = FALSE, cache = TRUE, cache.lazy=FALSE}
if(!is.null(params$output_file_summary)) write.csv(AUC_data, params$output_file_summary)
```
```{r echo=FALSE}
GRN_list <- list()
```

```{r echo = FALSE}
eval_package = FALSE
if("Epiregulon" %in% benchmarked_packages) eval_package = TRUE
```

```{r cache = TRUE, cache.lazy=FALSE, eval=eval_package}
weight.args <- list(clusters = as.vector(GeneExpressionMatrix[[params$cluster_column]]), regulon = regulon,
                    expMatrix = GeneExpressionMatrix, peakMatrix = PeakMatrix, peak_assay = "counts",
                    method = "corr", tf_re.merge = TRUE, exp_assay = "normalizedCounts")
# epiregulon workflow
regulon.w <- do.call(addWeights, weight.args)
GRN_list[["Epiregulon"]] <- regulon.w
if(params$save_results)
    saveRDS(regulon.w, file.path(save_path, "Epiregulon_reprogram.rds"))
```

```{r echo = FALSE}
eval_package = FALSE
if("Pando" %in% benchmarked_packages) eval_package = TRUE
```

```{r child = system.file("reports/Pando_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = FALSE, cache.lazy=FALSE, eval=eval_package}
```

```{r cache = TRUE, cache.lazy=FALSE}
Seurat_obj  <- Seurat::RenameCells(Seurat_obj, new.names = paste0("reprogram#", colnames(Seurat_obj))) 
GRN_list[["Pando"]] <- Seurat_obj 
```

```{r echo = FALSE}
eval_package = FALSE
if("cellOracle" %in% benchmarked_packages) eval_package = TRUE
```

```{r child = system.file("reports/cellOracle_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE, eval=eval_package}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["cellOracle"]] <- links
```

```{r echo = FALSE}
eval_package = FALSE
if("FigR" %in% benchmarked_packages) eval_package = TRUE
```

```{r child = system.file("reports/FigR_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE,message=FALSE, eval=eval_package}
```

```{r cache = TRUE, cache.lazy=FALSE}
GRN_list[["FigR"]] <- FigR_GRN
```

```{r echo = FALSE}
eval_package = FALSE
if("GRaNIE" %in% benchmarked_packages) eval_package = TRUE
```

```{r child = system.file("reports/GRaNIE_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE,message=FALSE, eval=eval_package}
```

```{r cache = TRUE, cache.lazy=FALSE, eval=eval_package}
GRN_list[["GRaNIE"]] <- GRN_connections.all
```

```{r echo = FALSE}
eval_package = FALSE
if("scenicplus" %in% benchmarked_packages) eval_package = TRUE
```

```{r child = system.file("reports/Scenic_plus_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy=FALSE, eval=eval_package}
```

```{r cache = TRUE, cache.lazy=FALSE, message=FALSE}
GeneExpressionMatrix <- GeneExpressionMatrix_copy
louvain_clusters$barcode = gsub("(.*)(___)(.*)", paste0("\\3","#","\\1"), louvain_clusters$barcode)
louvain_clusters$barcode <- gsub("\\.","-",louvain_clusters$barcode)
SummarizedExperiment::colData(GeneExpressionMatrix)$cluster_cellOracle <- as.character(louvain_clusters$cluster[match(louvain_clusters$barcode, colnames(GeneExpressionMatrix))])
matrices_list <- list()
for(i in seq_along(GRN_list)){
    GRN <- GRN_list[[i]]
    method <- names(GRN_list)[i]
    activity.matrix <- get_activity_matrix(GRN = GRN,
                                           method = method,
                                           geneExprMatrix.sce = GeneExpressionMatrix,
                                           tfs = c("GATA6", "NKX2-1", "FOXA2"),
                                           exp_assay = "normalizedCoutns")
  matrices_list[[i]] <- activity.matrix 
}

names(matrices_list) <- names(GRN_list)
if(exists("scenic_plus_res"))
    marices_list <- c(matrices_list, scenic_plus_res)
```

```{r cache = FALSE, cache.lazy=FALSE, message=FALSE}
for(i in seq_len(nrow(group_combinations))){
    tf <- group_combinations[i,"transcription_factor"]
    res <- plotDataFromActivity(matrices_list, tf=tf,
                     labels = GeneExpressionMatrix$hash_assignment,
                     positive_elements_label=group_combinations[i,"experimental_treatment"],
                     negative_elements_label=group_combinations[i,"control_treatment"],
                     GeneExpressionMatrix,
                     experimental_treatment = group_combinations[i,"experimental_treatment"],
                     title = sprintf("Reprogram-seq, induced %s overexpression\n%s activity", tf, tf))
    plot(res[[1]])
    if(params$save_results){
        ggsave(file.path(save_path, "plots", paste0("Reprogram-seq_by_package_", tf, ".pdf")), width = 6, height = 4)
        write.csv(res[[2]], file.path(save_path, paste0("Reprogram-seq_", tf, "_plot_data.csv")))
    }
}
```

