---
title: "Comparison of different methods for calculationg weights in epiregulon with the use of reprogram-seq dataset"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
params:
    regulon: NULL
    include_other_packages: FALSE
    treatments: list(GATA6 = NULL, NKX2.1 = NULL)
    output_file_sumary: NULL
    cluster_column: "Clusters"
    regulon_script: ""
    data_file_paths: ""
    temp_folder: ""
    conda_exe: ""
    virtual_env: ""
    work_dir: ""
    output_file: "Reprogram-seq.html"
---

```{r setup, include=TRUE, cache = TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```
# Introduction
This report presents comparison across all method options for weight computation which are available
in the `epiregulon` package. As a ground truth the reprogram-seq dataset is used. It was generated as a result of the studies in which cells were modified to overexpress a selected transcription factor. The results are
presented in the form of curves showing the relation between false and true positive rates.

# Regulon preparation
```{r echo = FALSE}
regulon <- params$regulon
eval_regulon = FALSE
if(is.null(params$regulon)) eval_regulon = TRUE
```
```{r child = system.file("reports/get_matrices_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, cache.lazy = FALSE}
```
```{r file = params$regulon_script_path, eval = FALSE}
```
```{r child = system.file("reports/pruned_regulon_reprogram-seq.Rmd", package = "epiregulon.benchmark"), eval = eval_regulon, echo = eval_regulon}
```

# Preparation of the data frame to be plotted

```{r cache = TRUE}
library(epiregulon.benchmark)
group_combinations <- expand.grid("", names(params$treatments), NA, "up", "")
colnames(group_combinations) <- c("cell_line", "experimental_treatment", "control_treatment", "effect_direction", "transcription_factor")
control_treatment <- c()
for(i in seq_along(params$treatments)){
    if(is.null(params$treatments[[i]]))
        control_treatment[i] <- get_complementary_names(names(params$treatments)[i], unique(GeneExpressionMatrix$hash_assignment))
    else
        control_treatment[i] <- paste0(params$treatments[[i]], collapse = "|")
}

group_combinations$control_treatment <- control_treatment
group_combinations$transcription_factor <- unlist(lapply(group_combinations$experimental_treatment, function(x) map_treatment_to_tf(x, regulon)))
plot_data <- data.frame()
```

```{r results = "hide", cache = TRUE, eval = params$include_other_packages}
arg.list <- list()
arg.list$peakMatrix = PeakMatrix
arg.list$peak_assay = "counts"
weight_clusters  <- as.list(setNames(rep(params$cluster_column, 3), c("corr", "MI", "lmfit")))
for (method in  c("corr", "MI", "lmfit","logFC","wilcoxon")){
    arg.list$method = method
    partial_data <- prepare_plot_data(regulon = regulon,
                                      weight.args = arg.list,
                                      group_combinations = group_combinations,
                                      geneExprMatrix.sce = GeneExpressionMatrix,
                                      weight_clusters = weight_clusters[[method]],
                                      activity_clusters = NA)
    plot_data <- rbind(plot_data, partial_data)
}

```

```{r cache = TRUE, eval = params$include_other_packages}
library(dplyr)
library(ggplot2)

ggplot(plot_data, aes(x=FPR, y=TPR))+
    geom_line()+
    facet_grid(treatment~method)+
    xlab("False positive rate")+
    ylab("True positive rate")+
    ggtitle("ROC curves for reprogram-seq dataset")


AUC_data <- plot_data %>% group_by(cell_line, treatment, method, tf) %>%
    group_map(function(x,y) cbind(y, data.frame(AUC = calculate_AUC(x$FPR, x$TPR))))
AUC_data <- do.call(rbind, AUC_data)

print(AUC_data)
```
```{r echo = FALSE}
if(!is.null(params$output_file_summary)) write.csv(AUC_data, params$output_file_summary)
```
```{r cache = TRUE, eval = params$include_other_packages}
GRN_list <- list()
weight.args <- list()
# epiregulon workflow
weight.args$clusters <- as.vector(GeneExpressionMatrix$Clusters)
weight.args$regulon <- pruned.regulon
weight.args$expMatrix <- GeneExpressionMatrix
weight.args$peakMatrix = PeakMatrix
weight.args$peak_assay = "counts"
weight.args$method = "corr"
regulon.w <- do.call(addWeights, weight.args)
GRN_list[["Epiregulon"]] <- regulon.w
```

```{r child = system.file("reports/Pando_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, eval = params$include_other_packages}
```

```{r cache = TRUE, eval = params$include_other_packages}
Seurat_obj  <- Seurat::RenameCells(Seurat_obj, new.names = paste0("reprogram#", colnames(Seurat_obj))) 
GRN_list[["Pando"]] <- Seurat_obj 
```


```{r child = system.file("reports/cellOracle_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, eval = params$include_other_packages}
```

```{r cache = TRUE, eval = params$include_other_packages}
GRN_list[["cellOracle"]] <- links
colData(GeneExpressionMatrix)$cluster_cellOracle <- louvain_clusters
```

```{r child = system.file("reports/FigR_reprogram-seq.Rmd", package = "epiregulon.benchmark"), cache = TRUE, eval = params$include_other_packages}
```

```{r cache = TRUE, eval = params$include_other_packages}
GRN_list[["FigR"]] <- FigR_GRN
```

```{r cache = TRUE, eval = params$include_other_packages}
saveRDS(GeneExpressionMatrix, "/gstore/scratch/u/wlodarct/GeneExpressionMatrix.rds")
saveRDS(GRN_list, "/gstore/scratch/u/wlodarct/GRN_list.rds")
matrices_list <- list()
for(i in seq_along(GRN_list)){
    GRN <- GRN_list[[i]]
    method <- names(GRN_list)[i]
    activity.matrix <- get_activity_matrix(GRN = GRN,
                                           method = method,
                                           geneExprMatrix.sce = GeneExpressionMatrix,
                                           tfs = c("GATA6", "NKX2-1", "FOXA2"))
  matrices_list[[i]] <- activity.matrix 
}

names(matrices_list) <- names(GRN_list)
```


```{r cache = TRUE, eval = params$include_other_packages}
AUC_table <- data.frame()
for (tf in c("GATA6", "NKX2-1", "FOXA2")){
    main_plot = TRUE
    positive_elements_label <- grep(tf, unique(GeneExpressionMatrix$hash_assignment), value = TRUE)
    negative_elements_label <- grep(tf, unique(GeneExpressionMatrix$hash_assignment), value = TRUE, invert = TRUE)
    for(i in seq_along(matrices_list)){
        activity.matrix <- matrices_list[[i]]
        if(!tf %in% rownames(activity.matrix)) next
        labels <- colData(GeneExpressionMatrix[,colnames(activity.matrix)])$hash_assignment
        if(main_plot) {
                AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                              tf = tf,
                              labels = labels,
                              add_plot = FALSE,
                              positive_elements_label = positive_elements_label,
                              negative_elements_label = negative_elements_label,
                              main = sprintf("Reprogram-seq dataset\n%s", tf))
                main_plot <- FALSE
        }
        else{
            AUC <- getResultsFromActivity(activity.matrix = activity.matrix,
                                  tf = tf,
                                  labels = labels,
                                  add_plot = TRUE,
                                  positive_elements_label = positive_elements_label,
                                  negative_elements_label = negative_elements_label)
        }
        AUC_table <- rbind(AUC_table, data.frame(AUC = AUC, tf = tf, method = names(matrices_list)[i]))
    }
}
print(AUC_table)
```

